\propertitle

### 2019-11-13 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-13T19:14:02+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

We want to stress the usefulness of making a quantified guess about the
activity of a larger brain area. Such a guess seems indeed to be the
primary idea behind recording a sample from that area. %  Second, it allows us
% to make comparisons across experimental sessions; such comparisons would me
% difficult or meaningless if made with the recorded samples, which generally
% comprise non-overlapping sets of neurons and differ in size.
There's also an important methodological reason: the question whether
correlations of some order are informationally sufficient cannot be
properly assessed at the sample level. When the difference in size between
sample and full network is too large, correlation sufficiency for the
sample implies the \emph{lack} of correlation sufficiency for the larger
network, and vice versa. This point is explained in detail in
\sect~\ref{sec:marginalization}.

@@

Maximum-entropy models have often been used to test whether a smaller part
of the sample data, for example time averages and second-order
correlations, is sufficient for quantifying our \dob\ about some aspects of
the neuronal activity of the sample. This use of maximum-entropy at the
sample level, however, has an important issue. We discuss this issue and
explain how our approach bypasses it in \sect~\ref{sec:marginalization}.
Moreover, our proposed model gives clearer answers to questions of
correlation sufficiency: we show this by applying it to the two data set
above addressing an example question of sufficiency, two moments versus
four moments.

@@

We first calculate the distribution by using six moments. This number
already provides almost as much information as the full frequency
distribution of the sample, and at the same time illustrates the use of the
approach in questions of statistical sufficiency \citep[cf. for
example][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}.
Figure~\ref{fig:plots_6mom} shows the resulting densities (that is,
distribution${}\times N$) for full-network sizes $N=n$, $N=1\,000$,
$N=5\,000$, $N=10\,000$. The case $N=n$ corresponds to applying the
maximum-entropy method at the sample level; it can be observed that with
six moments it reproduces almost exactly the measured frequency
distribution.
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/paper_stensola_6mom_distributionsN_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/paper_riehle_6mom_distributionsN_prior2.pdf}%
\caption{***}
\label{fig:plots_6mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb

The distribution for the full-network is sharper than the measured
frequency distribution for the sample; the sharper the larger $N$ is. Most
remarkably, it has two distinct low-activity modes for the first data set.
But also the second data set presents, upon closer inspection, a small
shoulder on the right of the mode, suggestive of two activity regimes. We
frame no hypotheses about the biological cause of these two modes (they
could stem from the presence of different kinds of cells or modules). These
features are clearly not present in the sample or in the maximum-entropy
distribution at the sample level. The application of the probability
calculus thus reveals interesting possible features of the full network.
\textcolor{white}{If you find this you can claim a postcard from us.}

@@

To illustrate how our approach can be applied to studies of sufficient
statistics, \fig~\ref{fig:plots_2_4mom} shows the results for two
constrained moments (equivalent to constraining means and correlations
only) and for four constrained moments, with $N=10\,000$. In either data
set we obtain two very distinct distributions. For the first data set in
particular, four moments lead to a bimodal distribution, whereas two to a
unimodal one, showing that two moments would not be sufficient
statistics. For comparison, the two distributions obtained applying
maximum-entropy \emph{at the sample level} are shown as an inset in the
plot for the first data set: their differences aren't so glaring as in the
full-network application.
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/paper2_stensola_twomom_distributions10000_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/paper_riehle_twomom_distributions10000_prior2.pdf}%
\caption{***}
\label{fig:plots_2_4mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb

@@

In the neurosciences the maximum-entropy method is often applied to
questions of sufficient statistics; for example: do pairwise correlations
in the neuronal activity provide the same information as the full frequency
distribution of activities? In this section we discuss how our proposed
application bears on this kind of questions. % For concreteness we use an
% example with specific statistics: first and second moments (equivalent to
% mean and correlations), versus first-to-fourth moments. But our discussion
% holds for any sets of statistics.

Let's make the question more precise first. We want to assess our degree of
belief about the frequencies of the activities of the sample, or about the
activity of the sample in a new time bin. For this assessment we should use
all measured data we have. But it sometimes happens that dropping part of
the data -- for example, the measured third- and higher-order moments --
leaves our assessment almost unchanged: the dropped data are
\emph{informationally irrelevant}, or almost so. The remaining data -- for
example, first and second moments -- are then \emph{informationally
  sufficient}. This informational quasi-sufficiency is interesting because
it may hint at peculiar biological  mechanisms or dependencies.

@@

In the neurosciences this check is often done with maximum-entropy
distributions constructed \emph{for the sample}, ignoring the full network
it's sampled from. The method introduced in the present work, however,
reveals an issue with this application at the sample level. The issue
doesn't come from the method itself but from a general fact of probability
theory. If a probability distribution has some sufficient statistics
\citep[\sect~4.5]{bernardoetal1994}{fortinietal2000}, then its marginals
\emph{cannot} have the same sufficient statistics, and vice versa, except
for trivial cases such as uniform distributions. This generally also holds
among different marginals: if a marginal has some sufficient statistics,
another won't have it. This impossibility is known in statistical
mechanics: marginals of Gibbs states aren't Gibbs states
\citep{maesetal1999}. Mathematically this impossibility translates into a
system of $n$ independent equations in $m$ unknowns with $n>m$
\citep[\sect~3.1]{portamanaetal2015}[\sect~3]{portamanaetal2018b}.

@@

The calculations and conclusions presented in the preceding sections depend
on the size of the full network, $N$. The full network can't be the whole
brain, of course; that would be silly. How large can $N$ be?

@@

This requirement delimits an area around the probe. Its number of neurons
depends on the animal species and on the brain region from which the
recording was made, as the neuron density can be very different. It's
practically impossible to specify the exact number of neurons, but its
order of magnitude is enough for making qualitative inferences. As we'll
discuss in the next section, even if the exact number were known the
maximum-entropy distribution ought still to be interpreted qualitatively.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-13T20:29:21+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal network\\ from subnetwork data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

@@

We want to stress the usefulness of making a quantified guess about the
activity of a larger brain area. Such a guess seems indeed to be the
primary idea behind recording a sample from that area. %  Second, it allows us
% to make comparisons across experimental sessions; such comparisons would me
% difficult or meaningless if made with the recorded samples, which generally
% comprise non-overlapping sets of neurons and differ in size.
There's also an important methodological reason. Maximum-entropy methods
have been used to assess the informational sufficiency of pairwise
correlations versus correlations of higher order \citep[see for
example][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}.
But when the difference in size between sample and full network is too
large, correlation sufficiency for the sample implies the \emph{lack} of
correlation sufficiency for the larger network, and vice versa, as will be
explained in detail in \sect~\ref{sec:marginalization}. Therefore,
maximum-entropy applications at the sample level can be deceptive for
questions of statistical sufficiency, and a full-network application is
more reliable. We show this point in \sect~\ref{sec:application} for a real
data set.

% Maximum-entropy models have often been used to test whether a smaller part
% of the sample data, for example time averages and second-order
% correlations, is sufficient for quantifying our \dob\ about some aspects of
% the neuronal activity of the sample. This use of maximum-entropy at the
% sample level, however, has an important issue. We discuss this issue and
% explain how our approach bypasses it in \sect~\ref{sec:marginalization}.
% Moreover, our proposed model gives clearer answers to questions of
% correlation sufficiency: we show this by applying it to the two data set
% above addressing an example question of sufficiency, two moments versus
% four moments.

@@

We first calculate the distribution by using six moments. This number
already provides almost as much information as the full frequency
distribution of the sample, and at the same time illustrates the use of the
approach in questions of statistical sufficiency.
Figure~\ref{fig:plots_6mom} shows the resulting densities (that is,
distribution${}\times N$) for four example values of full-network sizes:
$N=n$, $N=1\,000$, $N=5\,000$, $N=10\,000$. The case $N=n$ corresponds to
applying the maximum-entropy method at the sample level; it can be observed
that with six moments it reproduces almost exactly the measured frequency
distribution. We discuss the case of unknown $N$ in \sect~\ref{sec:N}.
\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/paper_stensola_6mom_distributionsN_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/paper_riehle_6mom_distributionsN_prior2.pdf}%
\caption{***}
\label{fig:plots_6mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb

The figure shows that the distribution for the full-network is less flat
than the measured frequency distribution for the sample; their difference
increases with $N$. Most remarkably, for the first data set the
distribution has two distinct low-activity modes. For the second data
the distribution presents a small shoulder on the right of the
mode.% , suggestive of two activity regimes. We frame no hypotheses about the
% biological cause of these two modes (they could stem from the presence of
% different kinds of cells or modules)
These features are clearly not present in the sample or in the
maximum-entropy distribution at the sample level. The application of the
probability calculus thus reveals interesting possible features of the full
network. \textcolor{white}{If you find this you can claim a postcard from
  us.}

@@

\begin{figure}[!p]
\centering
\includegraphics[width=\linewidth]{_scripts/paper2_stensola_twomom_distributions10000_prior2.pdf}%
\\[3em]%
\includegraphics[width=\linewidth]{_scripts/paper_riehle_twomom_distributions10000_prior2.pdf}%
\caption{***}
\label{fig:plots_2_4mom}
\end{figure}
%maxent_solutions_different_moments_prior2_6moments.nb
To illustrate how our approach can be applied to studies of sufficient
statistics, \fig~\ref{fig:plots_2_4mom} shows the full-network
distributions, for $N=10\,000$, obtained constraining the first two moments
(equivalent to constraining means and correlations) and the first four
moments; the frequency distribution of the sample is also shown for
comparison. In both data sets, the two-moment constrain leads to a
distribution quite different from the four-moment constrain. For the first
data set in particular, one distribution is bimodal, the other unimodal,
showing that two moments are not informationally sufficient.

It's useful to contrast the distributions for the full network with those
obtained applying the maximum-entropy method \emph{at the sample level}
with the same constraints, as traditionally done. The sample-level
distributions are shown as an inset in the plot for the first data set. We
see that constraining four moments doesn't lead to a distribution much
different from constraining only two. This would suggest that two moments
are informationally almost sufficient. This conclusion would be deceptive:
it obviously cannot be extended to the full network. We discuss this point
more in detail in the next section.

@@

As mentioned in the Introduction (see references there), in the
neurosciences the maximum-entropy method has also been applied to questions
of sufficient statistics. For example: do pairwise correlations in the
neuronal activity provide the same information as the full frequency
distribution of activities? In this section we discuss how our proposed
application bears on this kind of questions. % For concreteness we use an
% example with specific statistics: first and second moments (equivalent to
% mean and correlations), versus first-to-fourth moments. But our discussion
% holds for any sets of statistics.

The question must be made more precise first. We want to assess our degree
of belief about the frequencies of the activities of the sample, or about
the activity of the sample in a new time bin. For this assessment we should
use all measured data we have. But it sometimes happens that dropping part
of the data -- for example, the measured third- and higher-order moments --
leaves our assessment almost unchanged: the dropped data are
\emph{informationally irrelevant}, or almost so. The remaining data -- for
example, first and second moments -- are then \emph{informationally
  sufficient}. % This informational quasi-sufficiency is interesting because
% it may give clues about network organization and dynamics.

@@

In the neurosciences this check is often done with maximum-entropy
distributions constructed \emph{for the sample}, ignoring the full network
it's sampled from \citep[for
example][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}.
The method introduced in the present work, however, reveals an issue with
this application at the sample level. The issue doesn't come from the
method itself but from a general fact of probability theory. If a
probability distribution has some sufficient statistics
\citep[\sect~4.5]{bernardoetal1994}{fortinietal2000}, then its marginals
\emph{cannot} have the same sufficient statistics, and vice versa, except
for trivial cases such as uniform distributions. This generally also holds
among different marginals: if a marginal has some sufficient statistics,
another won't have it. This impossibility is known in statistical
mechanics: marginals of Gibbs states aren't Gibbs states
\citep{maesetal1999}. Mathematically this impossibility translates into a
system of $n$ independent equations in $m$ unknowns with $n>m$
\citep[\sect~3.1]{portamanaetal2015}[\sect~3]{portamanaetal2018b}.

@@

The calculations and conclusions presented in the preceding sections depend
on the size of the full network, $N$. The full network can't be the whole
brain, of course. How large should or can $N$ be in our formulae?

@@

This requirement delimits an area around the probe. Its number of neurons
depends on the animal species and on the brain region from which the
recording was made, as the neuron density can be very different. It's
practically impossible to specify the exact number of neurons, but its
order of magnitude is enough for making qualitative inferences. As
discussed in the next section, even if the exact number were known, the
maximum-entropy distribution ought to be interpreted qualitatively.
